Clustering time series based on probability distributions across temporal granularities

### 2021-11-18 main.Rmd
added id in table
minor corrections

There is an extensive body of literature focused on time series clustering related to smart meter data. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. <!--None of the $34$ articles chosen for their emphasis use Australian smart meter data.--> The most often used algorithm is $k$-means [@Rhodes2014-it]. 
$k$-means can be made to perform better by explicitly incorporating time series features such as correlation or cyclic patterns rather than performing it on raw data. To reduce dimensionality, several studies use principal component analysis (PCA) or factor analysis to pre-process smart-meter data before clustering [@Ndiaye2011-pf]. PCA eliminates correlation patterns and decreases feature space, but loses interpretability. Other algorithms utilized in the literature include $k$-means variants, hierarchical clustering, and greedy $k$-medoids. Time series data, such as smart meter data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics by first conducting a Fourier transformation, to convert data from time to frequency domain, followed by $k$-means to cluster by greatest frequency. @Motlagh2019-yj suggests that the time feature extraction is limited by the type of noisy, patchy, and unequal time series common in residential customers and addresses model-based clustering by transforming the series into other objects such as structure or set of parameters which can be more easily characterized and clustered. [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. @Melnykov2013-sp discusses how outliers, noisy observations and scattered observations can complicate estimating mixture model parameters and hence the partitions. None of these methods focuses on exploring heterogeneity in repetitive patterns based on the dynamics of multiple temporal dependencies using probability distributions, which forms the basis of the methodology reported here.

@@

Compute weighted pairwise distances $wpd$ [@Gupta2021-hakear] for all considered granularities for all observations. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction. It is estimated by the maximum pairwise distances between distributions across consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting patterns are expected, whereas a lower value would indicate otherwise.

@@

All of the common approaches can give contradictory suggestions for the optimal number of clusters, particularly when the data does not naturally break into groups, or in the presence of nuisance variables (no contribution to clustering) or nuisance observations (inlying outlying observations falling between clusters). There is no one best metric, which is perhaps a reason why metrics so many exist.

In this work, we have chosen to use $sindex$. It is a very simple, but effective metric. This is computed by averaging the smallest 10% of inter-cluster distances. It is relatively robust to nuisance observations. The value of $sindex$ always decreases, and sharp drops in value indicate candidates for the optimal number of clusters. The number of clusters corresponding to the value **before the drop** is the recommendation.

@@

To validate our clustering methods, have created several different data designs containing different granularity features. <!--to see where one method works better than the other and where they might give us the same outcome --or the effect of missing data <!---and trends on the proposed methods.--> There are three circular granularities $g1$, $g2$ and $g3$ with categories denoted by $\{g10,g11\}$, $\{g20, g21, g22\}$ and $\{g30, g31, g32, g33, g34\}$ and levels $n_{g_1}=2$, $n_{g_2}=3$ and $n_{g_3}=5$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by $\{0, 1, 2, \dots, 6\}$ or $\{Mon, Tue, \dots, Sun\}$. Here categories of $g1$, $g2$ and $g3$ are represented by $\{0, 1\}$, $\{0, 1, 2\}$ and $\{0, 1, 2, 3, 4\}$ respectively. A continuous measured variable $v$ of length $T$ indexed by $\{0, 1, \dots T-1\}$ is simulated such that it follows the structure across $g1$, $g2$ and $g3$. We constructed independent replications of all data designs $R = \{25, 250, 500\}$ to investigate if our proposed clustering method can discover distinct designs in small, medium, and big numbers of series. All designs employ $T=\{300, 1000, 5000\}$ sample sizes to evaluate small, medium, and large-sized series. Variations in method performance may be due to different jumps between categories. So a mean difference of $\mu = \{1, 2, 5\}$ between categories is considered. The performance of the approaches varies with the number of granularities which has interesting patterns across its categories. So three scenarios are considered to accommodate that. Table \ref{tab:range-parameter} shows the range of parameters considered for each scenario.

@@

```{r parcoord-sim, fig.cap = "Exploring the contribution of granularities in the clustering for scenarios S1, S2, S3, using parallel coordinate plots. Inter-cluster distances are displayed vertically. All three granularities $g1$, $g2$, and $g3$ have high inter-cluster distances for S1, suggesting all are important. In S2 g1 and in S3 both g1 and g2 have smaller inter-cluster distances, indicating that they did not contribute to clustering."}

@@

1. Choose a smaller subset of randomly selected $600$ customers with no implicit missing values for 2013.
2. Obtain $wpd$ for all cyclic granularities considered for these customers. It was found that $hod$ (hour-of-day), $moy$ (month-of-year) and $wnwd$ (weekend/weekday) are turning out to be significant for most customers. We use these three granularities while clustering.
3. Remove customers whose data for an entire category of $hod$, $moy$ or $wnwd$ is empty. For example, a customer who does not have data for an entire month is excluded because their monthly behavior cannot be analyzed.
4. Remove customers whose energy consumption is 0 in all deciles. These are the clients whose consumption is likely to remain essentially flat and with no intriguing repeated patterns that we are interested in studying.

@@

```{r opt_clusters-wpd}
```

```{r summary-group-wpd}
```

```{r summary-table-wpd}

@@

```{r summary-plot-wpd, fig.cap = "Summary plots for three (a) and five (b) clusters from WPD showing the $wpd$ values of each customers across $hod$, $moy$, and $wnwd$ through a parallel coordinate plot. Group P-1 and P-2 correspond to Q-1 and Q-2 respectively. Cluster P-3 splits into Q-3, Q-4 and Q-5. ", fig.height=6}
```

@@

The authors thank the ARC Centre of Excellence for Mathematical and Statistical Frontiers [(ACEMS)](https://acems.org.au/home) for supporting this research. Sayani Gupta was partially funded by [Data61 CSIRO](https://data61.csiro.au/) during her PhD. The Monash eResearch Centre and eSolutions-Study Support Services supported this research in part through the resource usage of the MonARCH HPC Cluster. The Github repository, [github.com/Sayani07/paper-gracsr](https://github.com/Sayani07/paper-gracsr), contains all materials required to reproduce this article and the code is also available online in the supplemental materials. This article was created with R [@R-language], `knitr` [@knitr, @R-knitr] and `rmarkdown` [@rmarkdown, @R-rmarkdown]. Graphics are produced with `ggplot2` [@Wickham2009pk] and `GGally` [@R-GGally].

>>>>

There is an extensive body of literature focused on time series clustering related to smart meter data. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. <!--None of the $34$ articles chosen for their emphasis use Australian smart meter data.--> The most often used algorithm is $k$-means [@Rhodes2014-it]. 
$k$-means can be made to perform better by explicitly incorporating time series features such as correlation or cyclic patterns rather than performing it on raw data. To reduce dimensionality, several studies use principal component analysis (PCA) or factor analysis to pre-process smart-meter data before clustering [@Ndiaye2011-pf]. PCA eliminates correlation patterns and decreases feature space, but loses interpretability. Other algorithms utilized in the literature include $k$-means variants, hierarchical clustering, and greedy $k$-medoids. Time series data, such as smart meter data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics by first conducting a Fourier transformation, to convert data from time to frequency domain, followed by $k$-means to cluster by greatest frequency. @Motlagh2019-yj suggests that the time feature extraction is limited by the type of noisy, patchy, and unequal time series common in residential customers and addresses model-based clustering by transforming the series into other objects such as structure or set of parameters which can be more easily characterized and clustered. @chicco2010renyi addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. @Melnykov2013-sp discusses how outliers, noisy observations and scattered observations can complicate estimating mixture model parameters and hence the partitions. None of these methods focuses on exploring heterogeneity in repetitive patterns based on the dynamics of multiple temporal dependencies using probability distributions, which forms the basis of the methodology reported here.

@@

We compute weighted pairwise distances $wpd$ [@Gupta2021-hakear] for all considered granularities for all observations. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction. It is estimated by the maximum pairwise distances between distributions across consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting patterns are expected, whereas a lower value would indicate otherwise.

@@

All of the common approaches can give contradictory suggestions for the optimal number of clusters, particularly when the data does not naturally break into groups, or in the presence of nuisance variables (no contribution to clustering) or nuisance observations (inlying outlying observations falling between clusters). There is no one best metric, which is perhaps a reason why so many metrics exist.

In this work, we have chosen to use $sindex$. It is a very simple, but effective metric. This is computed by averaging the smallest 10% of inter-cluster distances. It is relatively robust to nuisance observations. The value of $sindex$ always decreases, and sharp drops in value indicate candidates for the optimal number of clusters. The recommended number of clusters corresponding to the value *before the drop*.

@@

To validate our clustering methods, we have created several different data designs containing different granularity features. <!--to see where one method works better than the other and where they might give us the same outcome --or the effect of missing data <!---and trends on the proposed methods.--> There are three circular granularities $g1$, $g2$ and $g3$ with categories denoted by $\{g10,g11\}$, $\{g20, g21, g22\}$ and $\{g30, g31, g32, g33, g34\}$ and levels $n_{g_1}=2$, $n_{g_2}=3$ and $n_{g_3}=5$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by $\{0, 1, 2, \dots, 6\}$ or $\{Mon, Tue, \dots, Sun\}$. Here categories of $g1$, $g2$ and $g3$ are represented by $\{0, 1\}$, $\{0, 1, 2\}$ and $\{0, 1, 2, 3, 4\}$ respectively. A continuous measured variable $v$ of length $T$ indexed by $\{0, 1, \dots T-1\}$ is simulated such that it follows the structure across $g1$, $g2$ and $g3$. We constructed independent replications of all data designs $R = \{25, 250, 500\}$ to investigate if our proposed clustering method can discover distinct designs in small, medium, and big numbers of series. All designs employ $T=\{300, 1000, 5000\}$ sample sizes to evaluate small, medium, and large-sized series. Variations in method performance may be due to different jumps between categories. So a mean difference of $\mu = \{1, 2, 5\}$ between categories is considered. The performance of the approaches varies with the number of granularities which has interesting patterns across its categories. So three scenarios are considered to accommodate that. Table \ref{tab:range-parameter} shows the range of parameters considered for each scenario.

@@

```{r parcoord-sim, fig.cap = "Exploring the contribution of granularities in the clustering for scenarios S1, S2, S3, using parallel coordinate plots. Inter-cluster distances are displayed vertically. All three granularities $g1$, $g2$, and $g3$ have high inter-cluster distances for S1, suggesting all are important. In S2 $g1$ and in S3 both $g1$ and $g2$ have smaller inter-cluster distances, indicating that they did not contribute to clustering."}

@@

1. Choose a smaller subset of randomly selected $600$ customers with no implicit missing values for 2013.
2. Obtain $wpd$ for all cyclic granularities considered for these customers. It was found that $hod$ (hour-of-day), $moy$ (month-of-year) and $wnwd$ (weekend/weekday) are significant for most customers. We use these three granularities while clustering.
3. Remove customers whose data for an entire category of $hod$, $moy$ or $wnwd$ is empty. For example, a customer who does not have data for an entire month is excluded because their monthly behavior cannot be analyzed.
4. Remove customers whose energy consumption is 0 in all deciles. These are the clients whose consumption is likely to remain essentially flat and with no intriguing repeated patterns that we are interested in studying.

@@

@@

@@

The authors thank the ARC Centre of Excellence for Mathematical and Statistical Frontiers [(ACEMS)](https://acems.org.au/home) for supporting this research. Sayani Gupta was partially funded by [Data61 CSIRO](https://data61.csiro.au/) during her PhD. The Monash eResearch Centre and eSolutions-Study Support Services supported this research in part through the resource usage of the MonARCH HPC Cluster. The Github repository, [github.com/Sayani07/paper-gracsr](https://github.com/Sayani07/paper-gracsr), contains all materials required to reproduce this article and the code is also available online in the supplemental materials. This article was created with R [@R-language], `knitr` [@knitr; @R-knitr] and `rmarkdown` [@rmarkdown; @R-rmarkdown]. Graphics are produced with `ggplot2` [@Wickham2009pk] and `GGally` [@R-GGally].
